Avoiding the Global Sort: \break A Faster Contour Tree Algorithm

### 2017-01-04 paint-contour-tree.tex
small changes

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
It is convenient to think of $f$ as a simplicial complex sitting in $\RR^{d+1}$,
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components. 
The contour tree is a concise representation of these events.
Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a ball in $\RR^d$,
except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.

@@

We assume that the Morse up (or down) degree of any point
in $\MM$ is at most $2$, and that the total Morse degree
is at most $3$. This immediately implies analogous bounds
on the up and down-degrees in $\reeb(\MM)$. This
is standard in the literature
among others, though it can be challenging to enforce in 
practice. Theoretically, a manifold with points of higher Morse degree
can be converted to one with the degree condition through
a process of \emph{vertex unfolding} (refer to
Section 6.3 in~\cite{HaEd10}). This effectively converts
a multi-saddle to a set of ordinary saddles.
A detailed discussion is provided in the end
of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
 For 2D manifolds,
this conversion does not increase the surface complexity.
In higher dimensions, the unfolding can increase complexity if
$\MM$ has vertices of non-constant degree.

Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

We first give an informal description of the algorithm.

@@

\begin{compactenum}
    \item Construct a painting of $\MM$ by a descending BFS 
    from maxima. This partitions the edges (into various colored
    sets), and each vertex is incident to at most $2$ colors.
    \item For each color, construct a (binomial) max-heap
    indexed by height of all critical points touched by that color.
    \item Repeat until all heaps are empty:
    \begin{compactenum}
        \item Find a critical point $w$ that is at the top
        of all heaps containing $w$.
        \item Add $w$ to the join tree. 
        \item Delete $w$ from all these heaps, and merge them.
    \end{compactenum}
\end{compactenum}

@@

\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
\end{definition}

For a fast running time, we need to find ripe vertices
quickly. Thus, the formal algorithm requires a 
number of data structures, described below.

@@

For any critical point $v$ let $\ell_v$ denote the length of the longest directed path passing through $v$ in the join tree. 
As $|H_v|$ only counts vertices in a $v$ to root path, trivially $|H_v|\leq \ell_v$, immediately implying the analog of \Thm{main-corr} for join trees.
Note however that there is fair amount of slack in this argument as $|H_v|$ may be significantly smaller than $\ell_v$.  
This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{lb}.

@@

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complexes. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
% One of the critical aspects
% of the procedure is that the rain alternately flows downwards and upwards, 
% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
When we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
opposite direction. This is crucial to ensure a linear running time, and \Fig{order} shows how failing to switch 
directions can lead to a blow up in complexity. 
The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
% While one can implement $\rain$ this way, it is conceptually easier
% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

>>>>

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
It is convenient to think of $f$ as a simplicial complex sitting in $\RR^{d+1}$,
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components. 
The contour tree is a concise representation of these events \cite{kobps-ctsssit-97,csa-cctad-03}.
%Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
%For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a topological ball in $\RR^d$,
except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.

@@

% This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
% has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
% we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.

Boundary facets allow us to capture the resulting surface pieces after our algorithm makes a horizontal cut. 
The above definition is convenient, as any point within a given facet has a well-defined height, including the boundary facets.

@@

We assume that the Morse up (or down) degree of any point
in $\MM$ is at most $2$, and that the total Morse degree
is at most $3$. This immediately implies analogous bounds
on the up and down-degrees in $\reeb(\MM)$. This
is standard in the literature
among others, though it can be challenging to enforce in 
practice. Theoretically, a manifold with points of higher Morse degree
can be converted to one with the degree condition through
a process of \emph{vertex unfolding} (refer to
Section 6.3 in~\cite{HaEd10}). This effectively converts
a multi-saddle to a set of ordinary saddles.
%A detailed discussion is provided in the end
%of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
For 2D manifolds,
this conversion does not increase the surface complexity.
In higher dimensions, the unfolding can increase complexity if
$\MM$ has vertices of non-constant degree.

% Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
% via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
% Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

We first give an informal, idealized description of the algorithm. 
That is, in the actual algorithm the details differ slightly
and additional data structures are used to improve the running time.

@@

\begin{compactenum}
%     \item Construct a painting of $\MM$ by a descending BFS 
%     from maxima. This partitions the edges (into various colored
%     sets), and each vertex is incident to at most $2$ colors.
%     \item For each color, construct a (binomial) max-heap
%     indexed by height of all critical points touched by that color.
    \item Construct a painting of $\MM$ by a descending BFS from each maximum. 
    This partitions the edges into colored sets, 
    and assigns each vertex to the color classes of the edges that touch it.
    \item For each color, construct a (binomial) max-heap of the critical points touched by that color, 
          indexed by the heights of the critical points.
    \item Initialize the join tree to consists of a set of isolated vertices, one for each color class.\\
          (We now iteratively fill in the rest of the join tree from the leaves towards the root.)
    \item Repeat until all heaps are empty:
    \begin{compactenum}
        \item Find a critical point $w$ that is at the top of all heaps containing $w$.
        \item Add $w$ to the join tree by connecting it to the lowest vertex 
              of each component of the join tree that contains a maximum whose color matches one of the heaps containing $w$.
        \item Delete $w$ from all these heaps, and merge them.
    \end{compactenum}
\end{compactenum}

@@

\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, 
%$v$ is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
if $v$ is in heap of $c$ then it is the highest vertex in that heap (using notation from \Sec{struct}, if $v\in T(\rep(c))$, then it is the highest in $T(\rep(c))$).
%%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
\end{definition}

To efficiently implement the above, we need a number of data structures.  
Namely, binomial heaps for efficient merges, union-find to avoid explicit recoloring, 
and a stack to hold unripe (though ripening) vertices, which is used to effectively amortize the cost of 
the otherwise expensive operation of finding ripe vertices. 
Note we also later show it suffices to assign each vertex to only a single color from each adjacent up-star, 
rather than the colors of all the edges which touch it.

% For a fast running time, we need to find ripe vertices
% quickly. Thus, the formal algorithm requires a 
% number of data structures, described below.

@@

For any critical point $v$ let $\ell_v$ denote the length of the longest directed path passing through $v$ in the join tree. 
As $|H_v|$ only counts vertices in a $v$ to root path, trivially $|H_v|\leq \ell_v$, immediately implying the analog of \Thm{main-corr} for join trees.
Note however that there is fair amount of slack in this argument as $|H_v|$ may be significantly smaller than $\ell_v$.  
This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{pathDecomp}.

@@

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complexes, shown below as $\rain(x,\NN)$. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
% One of the critical aspects
% of the procedure is that the rain alternately flows downwards and upwards, 
% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
When we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
opposite direction. This is crucial to ensure linear running time, and \Fig{order} shows how failing to switch 
directions can lead to a blow up in complexity. 
The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
% While one can implement $\rain$ this way, it is conceptually easier
% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.
